---
title: "Analysis: Compounding invasion risk in freshwater systems"
format: html
execute:
  echo: true
  warning: false
  message: false
---

<!-- FILL IN: this file is the single source of truth for every number, table, and
figure that appears in index.qmd. Nothing gets computed in the manuscript file -->

## 1. Setup

```{r}
#| label: setup
# renv::restore()  # FILL IN once renv is initialized (see README)

library(tidyverse)
library(sf)
library(terra)
library(rgbif)
# library(biomod2)

# Bulk raw data (GBIF pulls, HydroATLAS, FutureStreams) lives on the external
# drive, not in the project folder, to keep the internal disk free. Override
# with the FIW_DATA_ROOT environment variable on another machine. Stop early
# with a clear message if the drive is missing -- otherwise the GBIF pull code
# would see an "empty" folder and start re-downloading everything.
data_root <- Sys.getenv(
  "FIW_DATA_ROOT",
  "/Volumes/Disk_OR/Freshwater_Invasion_Risk_data"
)
if (!dir.exists(data_root)) {
  stop("Data drive not found at: ", data_root,
       "\nPlug in the external drive (or set FIW_DATA_ROOT) before rendering.")
}
gbif_raw_dir <- file.path(data_root, "gbif_raw")
hydroatlas_dir <- file.path(data_root, "hydroatlas")
futurestreams_dir <- file.path(data_root, "futurestreams")
if (!dir.exists(gbif_raw_dir) || length(list.files(gbif_raw_dir, pattern = "\\.rds$")) == 0) {
  warning("gbif_raw is empty or missing at ", gbif_raw_dir,
          " -- the occurrence pull sections will re-download from GBIF.")
}

dir.create("data/derived", showWarnings = FALSE, recursive = TRUE)
dir.create("outputs/figures", showWarnings = FALSE, recursive = TRUE)
dir.create("outputs/tables", showWarnings = FALSE, recursive = TRUE)
```

## 1a. Candidate species expansion (SInAS-derived, 2026-09-05)

<!-- Decision (discussed with user 2026-09-05): the original 13-species
     shortlist (Section 2 below) was a hand-picked "dirty dozen"-style set.
     To make the compounding-risk null-model test (research_proposal.md
     Section 6 step 4) a genuine global test rather than a curated example,
     build the candidate list systematically from SInAS -- the alien-species
     database already planned as a data source (research_proposal.md
     Section 5) -- filtered to freshwater habitat, then screened by GBIF
     occurrence-record depth. Species below the depth threshold get dropped
     the same way the two "marginal" species were in Section 4, not silently
     kept. Source: SInAS_AlienSpeciesDB_2.5.csv, Zenodo DOI 10.5281/zenodo.10038256
     (Seebens et al. 2020), downloaded to data/raw/sinas/. -->

```{r}
#| label: sinas-freshwater-filter
sinas_raw <- read_delim(
  "data/raw/sinas/SInAS_AlienSpeciesDB_2.5.csv",
  delim = " ",
  quote = "\"",
  show_col_types = FALSE
)

# habitat is a ";"-joined multi-value field (e.g. "freshwater; terrestrial") --
# keep any row that includes freshwater at all, since a species doesn't need
# to be freshwater-exclusive to be a freshwater invader.
sinas_freshwater <- sinas_raw |>
  filter(str_detect(habitat, "freshwater"))

cat("Freshwater-habitat records in SInAS:", nrow(sinas_freshwater), "\n")
cat(
  "Distinct freshwater alien species (raw Taxon field):",
  n_distinct(sinas_freshwater$Taxon),
  "\n"
)

sinas_species_candidates <- sinas_freshwater |>
  distinct(Taxon, .keep_all = FALSE) |>
  rename(species = Taxon) |>
  arrange(species)

write_csv(
  sinas_species_candidates,
  "data/derived/sinas_freshwater_species_raw_list.csv"
)
```

```{r}
#| label: sinas-gbif-backbone-match
#| cache: true
# Batch-resolve all candidate names against the GBIF backbone in one call
# (name_backbone_checklist is vectorised -- orders of magnitude faster than
# looping name_backbone() per species, which is what Section 2's 13-species
# match used).
backbone_path <- "data/derived/sinas_gbif_backbone_match.csv"

if (!file.exists(backbone_path)) {
  # First attempt (2026-09-05) used the default bucket_size = 300 and failed
  # with "try lower bucket_size or larger sleep" -- 2,721 names in one async
  # job overwhelmed the endpoint. Smaller buckets + longer pauses fixed it.
  backbone_matches <- rgbif::name_backbone_checklist(
    sinas_species_candidates$species,
    bucket_size = 100,
    sleep = 3
  )
  backbone_matches$species_requested <- sinas_species_candidates$species
  write_csv(backbone_matches, backbone_path)
} else {
  backbone_matches <- read_csv(backbone_path, show_col_types = FALSE)
}

# Keep confident matches only (EXACT or close FUZZY), drop anything GBIF
# couldn't resolve to a usable taxon key.
sinas_matched <- backbone_matches |>
  filter(!is.na(usageKey), matchType %in% c("EXACT", "FUZZY")) |>
  distinct(usageKey, .keep_all = TRUE) # collapse synonyms mapping to the same accepted key

cat(
  "Resolved to a GBIF taxon key:",
  nrow(sinas_matched),
  "of",
  nrow(sinas_species_candidates),
  "candidate names\n"
)
```

## 1b. Taxonomic scope filter (decided with user, 2026-09-05)

<!-- The raw SInAS "freshwater habitat" tag is not restricted to
     aquatic-obligate species -- it also tags any species that USES
     freshwater habitat at all, which pulled in 203 bird species (including
     mallard, at 29.6 million GBIF occurrences from eBird), plus turtles,
     crocodilians, and mammals. Left in, these would dominate and invalidate
     the compounding-risk species pool. Decision (discussed with user,
     2026-09-05): keep fish, crayfish/crabs/shrimp (Malacostraca), bivalves,
     gastropods, and aquatic plants; exclude birds, mammals, amphibians, and
     reptiles entirely (not "flagged for later" -- fully out of scope). -->

```{r}
#| label: sinas-taxonomic-filter
# GBIF's backbone frequently leaves `class` blank for fish (a known quirk --
# confirmed 2026-09-05 by checking the project's own 13 shortlisted species,
# e.g. Cyprinus carpio resolves with class = NA despite being Actinopterygii),
# so fish have to be caught via order, not class, for the NA-class rows.
fish_orders <- c(
  "Cypriniformes",
  "Perciformes",
  "Siluriformes",
  "Salmoniformes",
  "Cyprinodontiformes",
  "Characiformes",
  "Atheriniformes",
  "Esociformes",
  "Osteoglossiformes",
  "Clupeiformes",
  "Acipenseriformes",
  "Synbranchiformes",
  "Tetraodontiformes",
  "Anguilliformes",
  "Mugiliformes",
  "Osmeriformes",
  "Gasterosteiformes",
  "Beloniformes",
  "Lepisosteiformes",
  "Scorpaeniformes",
  "Percopsiformes",
  "Gymnotiformes",
  "Polypteriformes",
  "Amiiformes",
  "Gadiformes",
  "Gonorynchiformes",
  "Pleuronectiformes",
  "Syngnathiformes"
)

is_fish <- with(
  sinas_matched,
  class %in%
    c("Actinopterygii", "Petromyzonti", "Dipneusti", "Elasmobranchii") |
    (phylum == "Chordata" & is.na(class) & order %in% fish_orders)
)
is_crayfish_crab <- sinas_matched$class == "Malacostraca"
is_bivalve <- sinas_matched$class == "Bivalvia"
is_gastropod <- sinas_matched$class == "Gastropoda"

# Plants: taxonomy alone can't separate obligate-aquatic species (water
# hyacinth, giant salvinia) from wetland-margin/terrestrial-tolerant weeds
# within the same plant classes. Proxy: keep a plant species only if at
# least one of its SInAS records is tagged habitat == "freshwater" *exactly*
# (not a combined "freshwater; terrestrial" tag) -- biases toward more
# aquatic-obligate species. Documented limitation, not a precise filter --
# worth a manual literature check on whichever plant species pass this and
# the occurrence-count threshold, before they go into the final shortlist.
plant_species_strict_freshwater <- sinas_freshwater |>
  filter(habitat == "freshwater") |>
  distinct(Taxon) |>
  pull(Taxon)
is_plant <- sinas_matched$kingdom == "Plantae" &
  sinas_matched$species_requested %in% plant_species_strict_freshwater

taxon_group <- case_when(
  is_fish ~ "Fish",
  is_crayfish_crab ~ "Crayfish/crab (Malacostraca)",
  is_bivalve ~ "Bivalve",
  is_gastropod ~ "Gastropod",
  is_plant ~ "Plant (strict freshwater habitat tag only)",
  TRUE ~ NA_character_
)

sinas_matched_scoped <- sinas_matched |>
  mutate(taxon_group = taxon_group) |>
  filter(!is.na(taxon_group))

write_csv(sinas_matched_scoped, "data/derived/sinas_matched_scoped.csv")

cat(
  "In taxonomic scope:",
  nrow(sinas_matched_scoped),
  "of",
  nrow(sinas_matched),
  "GBIF-resolved candidates\n"
)
count(sinas_matched_scoped, taxon_group, sort = TRUE)
```

```{r}
#| label: sinas-occurrence-counts
#| cache: true
# Lightweight per-species occurrence COUNT (not a full record pull) to
# screen candidates by data depth before deciding which get a full SDM.
# Checkpointed to CSV so this multi-thousand-call loop is resumable if
# interrupted -- do not re-query keys already recorded.
count_path <- "data/derived/sinas_gbif_occurrence_counts.csv"

existing_counts <- if (file.exists(count_path)) {
  read_csv(count_path, show_col_types = FALSE) |>
    mutate(usageKey = as.numeric(usageKey))
} else {
  tibble(
    usageKey = numeric(),
    species_requested = character(),
    n_occurrences = numeric()
  )
}

# usageKey type wasn't consistent between a fresh CSV-backed skeleton
# (integer) and what name_backbone_checklist actually returns -- force both
# sides to numeric before the filter/bind so this doesn't silently drop or
# fail to match already-queried keys. Query only the taxonomically-scoped
# set (Section 1b) -- no point spending API calls counting mallard ducks.
sinas_matched_scoped <- sinas_matched_scoped |>
  mutate(usageKey = as.numeric(usageKey))

keys_to_query <- sinas_matched_scoped |>
  filter(!usageKey %in% existing_counts$usageKey)

if (nrow(keys_to_query) > 0) {
  new_counts <- pmap_dfr(
    list(keys_to_query$usageKey, keys_to_query$species_requested),
    function(key, sp) {
      n <- tryCatch(
        rgbif::occ_count(
          taxonKey = key,
          hasCoordinate = TRUE,
          hasGeospatialIssue = FALSE
        ),
        error = function(e) NA_real_
      )
      tibble(
        usageKey = as.numeric(key),
        species_requested = sp,
        n_occurrences = as.numeric(n)
      )
    },
    .progress = TRUE
  )
  all_counts <- bind_rows(existing_counts, new_counts)
  write_csv(all_counts, count_path)
} else {
  all_counts <- existing_counts
}

cat(
  "Occurrence counts resolved for",
  nrow(all_counts),
  "species (all taxa queried to date --",
  "includes out-of-scope taxa from before the Section 1b filter existed)\n"
)

# Restrict to the taxonomically-scoped set (Section 1b) before this feeds
# any species-count decision -- all_counts/count_path is kept as a general
# cache of everything ever queried, not a pre-filtered analysis table.
scoped_counts <- all_counts |>
  semi_join(sinas_matched_scoped, by = "usageKey") |>
  left_join(
    sinas_matched_scoped |> select(usageKey, taxon_group),
    by = "usageKey"
  )

write_csv(scoped_counts, "data/derived/sinas_scoped_occurrence_counts.csv")
cat(
  "Occurrence counts within taxonomic scope:",
  nrow(scoped_counts),
  "species\n"
)
```

```{r}
#| label: sinas-threshold-summary
# How many candidate species clear common data-depth thresholds -- this is
# the number that should drive the final species-count decision, not a
# guess. Mirrors the two-marginal-species cutoff logic already used for the
# original 13-species shortlist (research_proposal.md Section 4).
threshold_summary <- tibble(threshold = c(100, 500, 1000, 5000, 10000)) |>
  mutate(
    n_species_at_or_above = map_int(
      threshold,
      ~ sum(scoped_counts$n_occurrences >= .x, na.rm = TRUE)
    )
  )

write_csv(threshold_summary, "data/derived/sinas_threshold_summary.csv")
threshold_summary
```

```{r}
#| label: sinas-threshold-by-group
# Same breakdown split by taxon group -- shows where the numbers actually
# come from (e.g. whether "500+" is mostly fish, or whether crayfish/
# bivalves/gastropods/plants contribute meaningfully once birds etc. are
# already excluded).
threshold_by_group <- scoped_counts |>
  group_by(taxon_group) |>
  summarise(
    n_total = n(),
    n_ge_500 = sum(n_occurrences >= 500, na.rm = TRUE),
    n_ge_1000 = sum(n_occurrences >= 1000, na.rm = TRUE),
    n_ge_5000 = sum(n_occurrences >= 5000, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(n_ge_1000))

write_csv(threshold_by_group, "data/derived/sinas_threshold_by_group.csv")
threshold_by_group
```

## 2. Species shortlist

<!-- FILL IN: load from data/raw/ once the GBIF-checked species table is exported
     as a CSV; see research_proposal.md Section 4 for the source list -->

```{r}
#| label: species-list
species_list <- tribble(
  ~group     , ~species                      , ~common_name                 , ~gbif_occurrences ,
  "Fish"     , "Cyprinus carpio"             , "Common carp"                ,            326799 ,
  "Fish"     , "Micropterus salmoides"       , "Largemouth bass"            ,            179621 ,
  "Crayfish" , "Pacifastacus leniusculus"    , "Signal crayfish"            ,            114713 ,
  "Fish"     , "Oreochromis niloticus"       , "Nile tilapia"               ,            101395 ,
  "Fish"     , "Gambusia affinis"            , "Mosquitofish"               ,             93064 ,
  "Crayfish" , "Procambarus clarkii"         , "Red swamp crayfish"         ,             87607 ,
  "Fish"     , "Neogobius melanostomus"      , "Round goby"                 ,             85983 ,
  "Fish"     , "Ameiurus nebulosus"          , "Brown bullhead catfish"     ,             51581 ,
  "Bivalve"  , "Dreissena polymorpha"        , "Zebra mussel"               ,             43797 ,
  "Plant"    , "Eichhornia crassipes"        , "Water hyacinth"             ,             21326 ,
  "Fish"     , "Hypophthalmichthys molitrix" , "Silver carp"                ,             14333 ,
  "Fish"     , "Clarias gariepinus"          , "African sharptooth catfish" ,             11459 ,
  "Fish"     , "Channa argus"                , "Northern snakehead"         ,              7956
)

write_csv(species_list, "data/derived/species_shortlist.csv")
species_list
```

## 3. Occurrence data acquisition & cleaning

<!-- Section 6, step 1 of the proposal. Raw pulls land in data/raw/gbif/ (one
     file per species) and are never modified in place; cleaned/combined
     output goes to data/derived/.

     IMPORTANT SCOPE NOTE (2026-09-05, first real pull): this uses rgbif's
     synchronous occ_search(), capped at occ_limit_per_species records per
     species. Several shortlisted species have 50,000-300,000+ raw GBIF
     records (research_proposal.md Section 4) -- pulling ALL of them requires
     GBIF's asynchronous Download API (a registered GBIF account + email
     confirmation + a wait for the export to build), which is a manual step
     outside this notebook. This first pass is a same-day, no-account,
     capped sample per species, sufficient to prototype the cleaning +
     predictor-extraction + SDM pipeline end to end (research_proposal.md
     Section 11, action 5). Re-run with the full Download-API export before
     final model fitting. -->

```{r}
#| label: occurrence-taxon-match
# Resolve each species name to its GBIF backbone taxon key first -- safer
# than matching on the free-text scientificName string, which can pull in
# synonyms/homonyms.
taxon_keys <- species_list |>
  mutate(
    kingdom_hint = if_else(group == "Plant", "Plantae", "Animalia"),
    backbone = map2(
      species,
      kingdom_hint,
      ~ rgbif::name_backbone(name = .x, kingdom = .y)
    )
  ) |>
  unnest_wider(backbone, names_sep = "_") |>
  select(
    group,
    species,
    common_name,
    gbif_occurrences,
    usageKey = backbone_usageKey,
    matchType = backbone_matchType,
    confidence = backbone_confidence
  )

write_csv(taxon_keys, "data/derived/species_taxon_keys.csv")
taxon_keys |> select(species, usageKey, matchType, confidence)
```

```{r}
#| label: occurrence-pull
#| cache: true
dir.create(gbif_raw_dir, showWarnings = FALSE, recursive = TRUE)

occ_limit_per_species <- 10000 # see scope note above

pull_one_species <- function(
  key,
  species_name,
  limit = occ_limit_per_species,
  max_attempts = 5
) {
  raw_path <- file.path(
    gbif_raw_dir,
    paste0(gsub(" ", "_", species_name), ".rds")
  )
  if (file.exists(raw_path)) {
    return(readRDS(raw_path))
  } # never re-pull/overwrite raw files

  # GBIF's API occasionally returns transient 503s under load -- retry with
  # exponential backoff rather than letting one flaky call kill the whole
  # loop (this happened on the first 13-species attempt, 2026-09-05).
  #
  # 2026-09-06 incident: the 353-species bulk pull hung for 11+ hours on a
  # single species -- a TCP connection to GBIF sat in CLOSE_WAIT with the R
  # process at 0% CPU, never erroring and never returning, so the retry
  # logic above never even triggered (there's a difference between "the
  # server returned an error" and "the connection never comes back", and
  # only the first one was handled). Added an explicit connect+overall
  # timeout via curlopts so a stalled connection surfaces as a timeout
  # error within ~90s and falls into the existing retry/backoff path,
  # instead of hanging indefinitely and burning 11GB of RAM.
  for (attempt in seq_len(max_attempts)) {
    res <- tryCatch(
      rgbif::occ_search(
        taxonKey = key,
        hasCoordinate = TRUE,
        hasGeospatialIssue = FALSE,
        limit = limit,
        curlopts = list(http_version = 2, connecttimeout = 30, timeout = 90)
      ),
      error = function(e) e
    )
    if (!inherits(res, "error")) {
      saveRDS(res, raw_path)
      Sys.sleep(1) # be polite to the GBIF API between species
      return(res)
    }
    message(sprintf(
      "  %s: attempt %d/%d failed (%s), retrying...",
      species_name,
      attempt,
      max_attempts,
      conditionMessage(res)
    ))
    Sys.sleep(5 * attempt) # 5s, 10s, 15s, 20s backoff
  }
  stop(sprintf(
    "Failed to pull %s after %d attempts",
    species_name,
    max_attempts
  ))
}

occ_raw <- pmap(
  list(taxon_keys$usageKey, taxon_keys$species),
  ~ pull_one_species(..1, ..2)
)
names(occ_raw) <- taxon_keys$species

occ_pull_summary <- tibble(
  species = names(occ_raw),
  n_pulled = map_int(occ_raw, ~ nrow(.x$data %||% tibble()))
)
write_csv(occ_pull_summary, "data/derived/gbif_pull_summary.csv")
occ_pull_summary
```

```{r}
#| label: occurrence-cleaning
# Combine, then apply minimal cleaning: keep coordinates + key metadata,
# drop exact duplicate lat/lon per species. Freshwater/inland filtering
# (removing marine/estuarine misassignments), full spatial thinning, and the
# SInAS cross-check are separate steps still to be added -- see the flagged
# next actions below.
occ_clean <- map_dfr(names(occ_raw), function(sp) {
  d <- occ_raw[[sp]]$data
  if (is.null(d) || nrow(d) == 0) {
    return(tibble())
  }
  d |>
    select(any_of(c(
      "species",
      "decimalLatitude",
      "decimalLongitude",
      "year",
      "country",
      "basisOfRecord",
      "coordinateUncertaintyInMeters"
    ))) |>
    mutate(species_requested = sp)
}) |>
  distinct(
    species_requested,
    decimalLatitude,
    decimalLongitude,
    .keep_all = TRUE
  )

write_csv(occ_clean, "data/derived/gbif_occurrences_clean.csv")
count(occ_clean, species_requested, sort = TRUE)
```

<!-- STILL TO DO before this feeds the SDM step (Section 6, step 1):
     1. Freshwater/inland filter -- remove marine/estuarine coordinate
        misassignments (a known contamination issue for the fish species).
     2. SInAS cross-check -- confirm "established" status per region rather
        than trusting every GBIF point as a naturalized occurrence.
     3. Spatial thinning -- reduce sampling-bias inflation (GBIF favours
        wealthy/well-surveyed countries).
     4. Re-pull via GBIF's Download API for species where the 10,000-record
        cap likely undersamples the true range (carp, bass, tilapia, etc. --
        see gbif_pull_summary.csv vs. the raw counts in
        research_proposal.md Section 4).

     DECISION (2026-09-07, confirmed with user): everything pulled via
     occ_search() in this notebook (both the 13-species prototype and the
     353-species bulk pull below) is prototyping/pipeline-validation data,
     NOT the final dataset for model fitting. The user already has a GBIF
     account and will re-pull the complete record set per species via
     GBIF's asynchronous Download API (no per-page cap, no pagination
     throttling) before final SDMs are fit for publication. This is not a
     scientific compromise: spatial thinning (item 3 above) discards most
     raw records as redundant regardless of the starting cap, so a capped
     sample is sufficient to build and test the full pipeline end-to-end
     now. The real risk of a cap is under-sampling a species' geographic
     range edges before thinning -- which is exactly why the final,
     pre-publication run needs the complete Download API export, not a
     capped sample. -->


## 3a. Occurrence pull for the expanded 353-species set (research_proposal.md Section 3b, 2026-09-06)
```{r}
#| label: bulk-species-target-list
scoped_counts <- read_csv(
  "data/derived/sinas_scoped_occurrence_counts.csv",
  show_col_types = FALSE
)

bulk_target_species <- scoped_counts |>
  filter(n_occurrences >= 5000) |>
  distinct(usageKey, species_requested, taxon_group)

write_csv(bulk_target_species, "data/derived/bulk_353_species_target_list.csv")
cat("Target species for this pull:", nrow(bulk_target_species), "\n")
count(bulk_target_species, taxon_group, sort = TRUE)
```

```{r}
#| label: bulk-occurrence-pull
#| cache: true
bulk_occ_limit <- 1000 # reduced from 5000 (2026-09-07): individual species were
# taking 3-5+ hours each (some pages of the ~17-page/species pagination at the
# 5000 cap appear to be badly throttled by GBIF after hundreds of sequential
# calls from this IP) -- 1000 records/species is still a standard, sufficient
# depth for SDM fitting (it's one of our own screening thresholds) and cuts
# per-species page count roughly 5x. Species already pulled at the 5000 cap
# keep their existing (better) sample; this only affects species not yet pulled.

bulk_pull_summary_path <- "data/derived/bulk_353_gbif_pull_summary.csv"
failed_species_path <- "data/derived/bulk_353_failed_species.csv"

# 2026-09-06/07 incident #2: pull_one_species() calls stop() after exhausting
# its retries, which is correct behaviour for a 13-species prototype but
# fatal for an unattended 353-species job -- one permanently-failing species
# (Esox masquinongy, muskellunge) killed the entire pmap() and the process
# sat idle for hours afterward with no indication anything was wrong beyond
# a dead R process. Wrap each call so one bad species is logged and skipped,
# never fatal to the batch. Also drop max_attempts for the bulk run (3
# instead of the default 5) -- sinking 5 full retries with growing backoff
# into a species that's already failing is wasted wall-clock time at this
# scale; log it and move on rather than stall the whole run on one name.
safe_pull_one_species <- function(key, species_name) {
  tryCatch(
    pull_one_species(
      key,
      species_name,
      limit = bulk_occ_limit,
      max_attempts = 3
    ),
    error = function(e) {
      message(sprintf(
        "PERMANENTLY FAILED: %s (%s) -- logged, continuing",
        species_name,
        conditionMessage(e)
      ))
      write_csv(
        tibble(
          species = species_name,
          usageKey = key,
          error = conditionMessage(e),
          failed_at = as.character(Sys.time())
        ),
        failed_species_path,
        append = file.exists(failed_species_path)
      )
      NULL
    }
  )
}

bulk_occ_raw <- pmap(
  list(bulk_target_species$usageKey, bulk_target_species$species_requested),
  ~ safe_pull_one_species(..1, ..2)
)
names(bulk_occ_raw) <- bulk_target_species$species_requested

bulk_pull_summary <- tibble(
  species = names(bulk_occ_raw),
  n_pulled = map_int(
    bulk_occ_raw,
    ~ if (is.null(.x)) NA_integer_ else nrow(.x$data %||% tibble())
  )
)
write_csv(bulk_pull_summary, bulk_pull_summary_path)
cat(
  "Species attempted:",
  nrow(bulk_pull_summary),
  "| succeeded:",
  sum(!is.na(bulk_pull_summary$n_pulled)),
  "| permanently failed:",
  sum(is.na(bulk_pull_summary$n_pulled)),
  "\n"
)
summary(bulk_pull_summary$n_pulled)
```

```{r}
#| label: bulk-occurrence-cleaning
# Same minimal cleaning as the 13-species prototype (Section 3) -- combine,
# keep coordinates + key metadata, drop exact duplicate lat/lon per species.
# The same "STILL TO DO" list above (freshwater/inland filter, SInAS
# established-status cross-check, spatial thinning) applies here too, at
# the full 353-species scale.
bulk_occ_clean <- map_dfr(names(bulk_occ_raw), function(sp) {
  d <- bulk_occ_raw[[sp]]$data
  if (is.null(d) || nrow(d) == 0) {
    return(tibble())
  }
  d |>
    select(any_of(c(
      "species",
      "decimalLatitude",
      "decimalLongitude",
      "year",
      "country",
      "basisOfRecord",
      "coordinateUncertaintyInMeters"
    ))) |>
    mutate(species_requested = sp)
}) |>
  distinct(
    species_requested,
    decimalLatitude,
    decimalLongitude,
    .keep_all = TRUE
  )

write_csv(bulk_occ_clean, "data/derived/bulk_353_gbif_occurrences_clean.csv")
cat(
  "Total cleaned occurrence records across the 353-species set:",
  nrow(bulk_occ_clean),
  "\n"
)
```

## 4. Predictor assembly (HydroATLAS + FutureStreams)

<!-- FILL IN: extract HydroATLAS static variables + FutureStreams derived
     bioclim-style variables at occurrence points, snapped to the river
     network. Section 6, step 2 -->

```{r}
#| label: predictor-assembly
#| eval: false
```

## 5. Individual species SDMs

<!-- FILL IN: biomod2 ensemble models per species, current + future
     projections per RCP/warming level. Section 6, step 3 -->

```{r}
#| label: species-sdms
#| eval: false
#| cache: true
```

## 6. Compounding-risk metric

<!-- FILL IN: overlay predicted suitable-range rasters, build overlap/
     stacking index, spatial randomization null-model test. Section 6, step 4 -->

```{r}
#| label: compounding-risk
#| eval: false
```

## 7. Validation against real co-occurrence records

<!-- FILL IN: cross-reference high-overlap cells against documented
     multi-species co-occurrence (GBIF/SInAS/CABI) to flag "primed but not
     yet hit" zones. Section 6, step 5 -->

```{r}
#| label: validation
#| eval: false
```

## 8. Biodiversity-value cross-reference

<!-- FILL IN: overlay against WWF Freshwater Ecoregions endemism data.
     Section 6, step 6 -->

```{r}
#| label: biodiversity-overlay
#| eval: false
```

## 9. Native-refugia overlay (extension)

<!-- FILL IN: compare against Wang et al. (2026)-style thermal-refugia
     zones using the same FutureStreams data. Section 6, step 7 -->

```{r}
#| label: refugia-overlay
#| eval: false
```

## 10. Sensitivity / robustness checks

<!-- FILL IN: vary overlap threshold, alternative null models, drop
     highest-occurrence-count species to check they aren't driving results.
     Section 6, step 8 -->

```{r}
#| label: robustness
#| eval: false
```

## 11. Figures

```{r}
#| label: fig-placeholder
#| eval: false
# ggsave("outputs/figures/fig1_compounding_risk_map.png", plot, width = 8, height = 6, dpi = 300)
```

## 12. Session info

```{r}
#| label: session-info
sessionInfo()
```
