---
title: "Global compounding risk from co-occurring freshwater invasive species: a predictive test of invasional meltdown"
author:
  - name: "FILL IN: Your Name"
    affiliation: "FILL IN: Institution"
    email: "FILL IN"
date: today
format:
  pdf:
    documentclass: article
    linestretch: 2
  docx:
    default: true
  html:
    toc: true
bibliography: references/references.bib
csl: references/nature.csl
---

```{r}
#| label: setup
#| echo: false
library(tidyverse)
species_list <- read_csv("data/derived/species_shortlist.csv", show_col_types = FALSE)
```

## Abstract

<!-- FILL IN: 150-250 words once results exist. Draft skeleton below. -->

Invasional meltdown — the hypothesis that non-native species facilitate one
another's establishment and impact — has been supported almost exclusively
by single-site case studies over 25 years. Here we provide, to our
knowledge, the first global, spatially-explicit, predictive test of the
hypothesis in freshwater systems, combining species distribution models for
`r nrow(species_list)` invasive freshwater taxa with river-network-aware
environmental data. FILL IN: headline result once analysis is complete.

## Introduction

Invasional meltdown — the hypothesis that non-native species facilitate one
another's establishment and impact [@Simberloff1999] — is a 25-year-old,
heavily cited concept. Despite this, the evidence base is built almost
entirely from single-site case studies or theoretical and review papers;
there is no global, predictive, spatially-explicit test of the hypothesis.

The most current authoritative synthesis [@Ricciardi2025] explicitly names
multi-species biotic-interaction effects as an open, unaddressed gap in
current invasion modeling. A systematic review of 150 invasional-meltdown
studies confirmed broad support but with biases and gaps [@Braga2018], and a
large IPBES-linked consortium assessment found that freshwater systems carry
among the largest documented knowledge gaps in global alien-species status
and trend assessment [@Seebens2025].

Species distribution modeling (SDM) of invasive species is mature and
well-established [@Early2016], but no study has used it to explicitly map
where multiple independently-introduced invaders' predicted ranges overlap,
as a quantitative signal of compounding or meltdown risk.

**The gap, stated plainly:** nobody has combined (a) global SDM/occurrence-based
range prediction, (b) multiple co-occurring freshwater invaders across taxa,
and (c) explicit cross-referencing against real-world co-occurrence records
and biodiversity value, to predict where the *next* invasional meltdown is
most likely — before it shows up in the observational record.

### Research questions

1. What is the current (and projected future) climatically/hydrologically
   suitable range of each shortlisted invasive freshwater species?
2. Where do multiple species' predicted suitable ranges spatially
   co-occur, more than expected under a null/random model?
3. Of the high-compounding-risk zones identified, which do not yet show
   documented multi-species co-occurrence in the real occurrence record —
   the "primed but not yet hit" zones?
4. Do those emerging zones disproportionately overlap freshwater
   biodiversity hotspots or high-endemism areas?
5. (Extension) Do projected future refugia for native fish coincide with
   zones of increasing invader-overlap suitability under the same climate
   forcing [@Wang2026]?

## Methods

### Species and occurrence data

`r nrow(species_list)` species spanning fish, crayfish, and bivalves across
five or more source regions were selected from the GBIF taxonomic backbone
(Table 1; see `analysis.qmd` for the full acquisition and cleaning pipeline).

```{r}
#| label: tbl-species
#| echo: false
#| tbl-cap: "Shortlisted invasive freshwater species."
species_list |> select(group, species, common_name, gbif_occurrences)
```

### Predictor variables

<!-- FILL IN: HydroATLAS + FutureStreams predictor description once the
     pipeline in analysis.qmd is finalized -->

### Species distribution modeling

<!-- FILL IN: ensemble SDM approach (biomod2), validation, projections -->

### Compounding-risk metric

<!-- FILL IN: overlap index definition and null-model test -->

### Validation and biodiversity overlay

<!-- FILL IN -->

## Results

<!-- FILL IN: populate once analysis.qmd produces outputs; reference
     outputs/figures/ and outputs/tables/ only, do not compute here -->

## Discussion

<!-- FILL IN -->

## Data availability

<!-- FILL IN: GBIF/SInAS/HydroATLAS/FutureStreams accession details; code
     and derived data archived at Zenodo upon submission (DOI: FILL IN) -->

## References

::: {#refs}
:::
